Bayesian calibration of a hybrid nuclear collision model\\ using $\boldsymbol p$-Pb and Pb-Pb data at energies available\\ at the CERN Large Hadron Collider

### 2019-10-14 paper/paper.tex
Begin implementing referee edits.

\title{
  Bayesian calibration of a hybrid nuclear collision model\\
  using $\boldsymbol p$-Pb and Pb-Pb data from the LHC
}

@@

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two nucleons collide inelastically at impact parameter $\bv$.
In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll}(\bv) = 1 - \exp[-\sigma_\mathrm{eff}\, T_{nn}(\bv)],
\end{equation}
where $\sigma_\mathrm{eff}$ is the effective cross section for inelastic collisions between each pair of constituents, determined by fitting the inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int d^2b\, P_\mathrm{coll}(\bv)
\end{equation}
at the specified collision energy $\sqrts$.

@@

The GPs are essentially fancy interpolators applied to the model's training points and PCA transformed observables.
Each GP reports a mean value $z(\xv)$ as well as an estimated error $\delta z(\xv)$ which accounts for statistical noise in the training data and interpolation error between the design points.
Once the GPs are trained, we can predict the observables $\ym$ at parameter point $\xv$ by transforming the vector of principal components
\begin{equation}
  \mathbf{z}(\xv) = [z_1(\xv), z_2(\xv), \dots, z_k(\xv)]
\end{equation}
\emph{back} to physical space.
Similarly, we can construct the covariance matrix of the observables in PCA space
\begin{equation}
  \cov(z_i, z_j) = \diag[(\delta z_{1})^2, (\delta z_2)^2, \dots, (\delta z_k)^2 ],
\end{equation}
and transform it back to physical space as well to obtain the covariance matrix $\Sigmam$ of the model observables at parameter point $\xv$.

@@

Conversely, the calibrated (posterior sampled) emulator predictions (bottom row of Figs.~\ref{fig:obs_pbpb} and \ref{fig:obs_ppb}) are far better constrained and nicely track the experimental data points.
We emphasize here that the posterior parameter values are obtained from a \emph{simultaneous} calibration to $p$-Pb and Pb-Pb data, and thus they are self-consistent between the two systems.
The spread in the posterior samples reflects different sources of model and experimental uncertainty as well as tension in the optimal fit parameters which describe each observable.
We demonstrate later in the text that a single set of model parameters well describes all of the calibration data, and thus we believe that much of the spread in the posterior samples is uncertainty contributed by our emulator.
We also note that although the $p$-Pb posterior samples have a somewhat larger spread than the Pb-Pb samples, the percentage uncertainty of the $p$-Pb emulator is similar to that of the Pb-Pb emulator, and thus the difference is likely due to the larger variance of the $p$-Pb training data.
The uncertainty in the posterior distribution could thus be improved by running the calibration with more design points or with a narrower range of parameter values to increase the density of the training points and reduce interpolation uncertainty.

@@

@@

This is perhaps the single largest difference between our work and the conclusions of recent saturation-based calculations which constrained the event-by-event fluctuations of the proton using a color dipole picture of vector meson production \cite{Mantysaari:2016ykx, Mantysaari:2016jaz}.
Those studies find that the measured coherent and incoherent $J/\Psi$ spectra at HERA prefer a compact gluon distribution inside each nucleon, with a Gaussian width $w_g \approx 0.4$~fm which is roughly \emph{half} the Gaussian width preferred by our analysis.
Evidently, it may be necessary to place an informative prior on our nucleon substructure parameters in order to resolve the apparent tension between our parameter values and those needed to describe DIS measurements at HERA.
Alternatively, it is also possible that the fluctuations probed by coherent and incoherent $J/\Psi$ production are different than those probed by minimum-bias particle production.

@@

@@

The global agreement of the MAP model calculations with the experimental data is very good.
The largest tension is observed in the two-particle cumulants $\vnk{2}{2}$ and $\vnk{3}{2}$ of the $p$-Pb system, although even that tension is only about 10--15\%.
Quite remarkably, the model perfectly describes the shape of the $p$-Pb and Pb-Pb two-particle correlations which is strong evidence that these correlations are hydrodynamic in origin.
Moreover, we obtain an excellent description of the \mbox{$p$-Pb} mean $p_T$, although this fit is somewhat less meaningful since we are unable to calibrate on the Pb-Pb mean $p_T$ simultaneously (data is not yet available).
Additionally, the model provides a simultaneous description of the \mbox{$p$-Pb} and Pb-Pb charged-particle yields using a single energy deposition parameter $p=0$.
This is the \emph{exact same} generalized mean $p$-value supported by multiple previous studies \cite{Moreland:2014oya, Bernhard:2016tnd, Ke:2016jrd, Bernhard:2018hnz}.
Evidently, this scaling continues to hold for initial conditions with sizable nucleon substructure.

@@

Lastly, we compute the symmetric cumulants $\SC(m,n)$ for the Pb-Pb collision system at $\sqrts=5.02$~TeV which quantify correlations between event-by-event fluctuations of the flow harmonics of different order \cite{Bilandzic:2013kga, ALICE:2016kpq}
\begin{align}
  \SC(m, n) &= \langle\langle \cos[m(\phi_1 - \phi_3) + n(\phi_2-\phi_4)]\rangle\rangle \nonumber \\
  \nonumber &- \langle\langle\cos[m(\phi_1-\phi_2)]\rangle\rangle\langle\langle\cos[n(\phi_1-\phi_2)]\rangle\rangle \label{eq:scmn}\\
  &\approx \langle v_m^2 v_n^2 \rangle - \langle v_m^2\rangle\langle v_n^2\rangle.
\end{align}
We show these model predictions in Fig.~\ref{fig:flow_corr} along with the \emph{normalized} symmetric cumulants
\begin{equation}
  \NSC(m,n) = \SC(m,n)/\langle v_m^2\rangle\langle v_n^2\rangle,
\end{equation}

@@

\noindent which are expected to be less sensitive to the medium response and more sensitive to the properties of the initial state.
The solid lines are the MAP estimate of the present study, and the dashed lines are the MAP estimate of Ref.~\cite{Bernhard:2018hnz} which did not include nucleon substructure and was calibrated on Pb-Pb collisions at $\sqrts=2.76$ and 5.02~TeV.
We observe that the gap between $\SC(3,2)$ and $\SC(4,2)$ is generally wider in the present analysis than in Ref.~\cite{Bernhard:2018hnz}, as is the gap between the normalized symmetric cumulants $\NSC(3,2)$ and $\NSC(4,2)$.

>>>>

\title{
  Bayesian calibration of a hybrid nuclear collision model\\
  using $\boldsymbol p$-Pb and Pb-Pb data at energies available\\
  at the CERN Large Hadron Collider
}

@@

Within a picture of independent pairwise collisions between the constituents, a Glauber model model may be used to calculate the probability $P_\mathrm{coll}$ that the two nucleons collide inelastically at impact parameter $\bv$.
In the limit when the number of constituents is large, it yields the particularly simple form
\begin{equation}
  \label{eq:pcoll}
  P_\mathrm{coll}(\bv) = 1 - \exp[-\sigma_\mathrm{eff}\, T_{nn}(\bv)],
\end{equation}
where $\sigma_\mathrm{eff}$ is the effective cross section for inelastic collisions between each pair of constituents, determined by fitting the inelastic proton-proton cross section
\begin{equation}
  \label{eq:sigma_nn}
  \sigma_{pp}^\mathrm{inel} = \int d^2b\, P_\mathrm{coll}(\bv)
\end{equation}
at the specified collision energy $\sqrts$.
We tune $\sigma_\mathrm{eff}$ in the present work to fit the experimental inelastic nucleon-nucleon cross section $\sigma_\text{NN}^\text{inel}= 7.0$~fm$^2$ at $\sqrts=5.02$~TeV for comparison with our chosen datasets \cite{ALICE:2012xs}.
The resulting \trento\ inelastic nucleon-nucleon cross section agrees with the experimental value to better than 2\% accuracy, as verified by one of the model's standard unit tests.

@@

The GPs are essentially fancy interpolators applied to the model's training points and PCA transformed observables.
Each GP reports a mean value $z(\xv)$ as well as an estimated error $\delta z(\xv)$ which accounts for statistical noise in the training data and interpolation error between the design points.
Once the GPs are trained, we can predict the observables $\ym$ at parameter point $\xv$ by transforming the vector of principal components
\begin{equation}
  \mathbf{z}(\xv) = [z_1(\xv), z_2(\xv), \dots, z_k(\xv)]
\end{equation}
\emph{back} to physical space.

Similarly, we can construct the covariance matrix of the observables in PCA space
\begin{equation}
  \cov(z_i, z_j) = \diag[(\delta z_{1})^2, (\delta z_2)^2, \dots, (\delta z_k)^2 ],
\end{equation}
and transform it back to physical space as well to obtain the covariance matrix $\Sigmam$ of the model observables at parameter point $\xv$.

@@

Conversely, the calibrated (posterior sampled) emulator predictions (bottom row of Figs.~\ref{fig:obs_pbpb} and \ref{fig:obs_ppb}) are far better constrained and nicely track the experimental data points.
We emphasize here that the posterior parameter values are obtained from a \emph{simultaneous} calibration to $p$-Pb and Pb-Pb data, and thus they are self-consistent between the two systems.
The spread in the posterior samples reflects different sources of model and experimental uncertainty as well as tension in the optimal fit parameters which describe each observable.
We demonstrate later in the text that a single set of model parameters well describes all of the calibration data, and thus we believe that much of the spread in the posterior samples is uncertainty contributed by our emulator.

\noindent
We also note that although the $p$-Pb posterior samples have a somewhat larger spread than the Pb-Pb samples, the percentage uncertainty of the $p$-Pb emulator is similar to that of the Pb-Pb emulator, and thus the difference is likely due to the larger variance of the $p$-Pb training data.
The uncertainty in the posterior distribution could thus be improved by running the calibration with more design points or with a narrower range of parameter values to increase the density of the training points and reduce interpolation uncertainty.

@@

This is perhaps the single largest difference between our work and the conclusions of recent saturation-based calculations which constrained the event-by-event fluctuations of the proton using a color dipole picture of vector meson production \cite{Mantysaari:2016ykx, Mantysaari:2016jaz}.
Those studies find that the measured coherent and incoherent $J/\Psi$ spectra at HERA prefer a compact gluon distribution inside each nucleon, with a Gaussian width $w_g \approx 0.4$~fm which is roughly \emph{half} the Gaussian width preferred by our analysis.
Evidently, it may be necessary to place an informative prior on our nucleon substructure parameters in order to resolve the apparent tension between our parameter values and those needed to describe DIS measurements at HERA.

\begin{figure}[b]
  \includegraphics{inelasticity_profile}
  \caption{
    \label{fig:inelasticity_profile}
    Glauber model estimates of the normalized proton-proton inelasticity density $G_\mathrm{inel}(b) = d^2\sigma_\mathrm{inel}/d^2b$ at two different LHC beam energies.
    Blue line: A multi-source Glauber model fit to LHC data at $\sqrts=7$~TeV \cite{Albacete:2016pmp}.
    Orange line: \trento\ model predictions at $\sqrts=5.02$~TeV using median parameters from the present study's Bayesian posterior.
  }
\end{figure}

Additional constraints on the proton size and shape are provided by the proton-proton inelasticity density $G_\mathrm{inel}(b) = d^2\sigma_\mathrm{inel}(b)/d^2b$, which measures the proton-proton inelastic collision probability as a function of its impact parameter.
We compare, in Fig~\ref{fig:inelasticity_profile}, the normalized proton-proton inelasticity density $G_\mathrm{inel}(b) / G_\mathrm{inel}(0)$ at $\sqrts=5.02$~TeV predicted by \trento\ using the present study's posterior median parameters with a calculation from reference \cite{Albacete:2016pmp}, which extracted the proton-proton inelasticity density from LHC data at $\sqrts=7$~TeV using a multi-source Glauber model \cite{Albacete:2016pmp}.
Our Bayesian median estimate (orange line) prefers a larger width for the proton-proton normalized inelasticity density compared to that of reference \cite{Albacete:2016pmp}, which is \emph{opposite} the trend one would expect based on the differences in beam energies alone.
This suggests that our posterior estimate for the proton radius is somewhat oversized.
Nevertheless, it is fascinating that the present Bayesian estimate is as close as it is, given that the model is fit to quantities which are not typically used to extract the proton-proton inelasticity density.

@@

@@

The global agreement of the MAP model calculations with the experimental data is generally quite good.
The largest tension is observed in the two-particle cumulants $\vnk{2}{2}$ and $\vnk{3}{2}$ of the $p$-Pb system, although even that tension is only about 10--15\%.
Quite remarkably, the model perfectly describes the shape of the $p$-Pb and Pb-Pb two-particle correlations which is strong evidence that these correlations are hydrodynamic in origin.

@@

Moreover, we obtain an excellent description of the \mbox{$p$-Pb} mean $p_T$, although this fit is somewhat less meaningful since we are unable to calibrate on the Pb-Pb mean $p_T$ simultaneously (data is not yet available).
Additionally, the model provides a simultaneous description of the \mbox{$p$-Pb} and Pb-Pb charged-particle yields using a single energy deposition parameter $p=0$.
This is the \emph{exact same} generalized mean $p$-value supported by multiple previous studies \cite{Moreland:2014oya, Bernhard:2016tnd, Ke:2016jrd, Bernhard:2018hnz}.
Evidently, this scaling continues to hold for initial conditions with sizable nucleon substructure.

@@

Lastly, we compute the symmetric cumulants $\SC(m,n)$ for the Pb-Pb collision system at $\sqrts=5.02$~TeV which quantify correlations between event-by-event fluctuations of the flow harmonics of different order \cite{Bilandzic:2013kga, ALICE:2016kpq}
\begin{align}
  \SC(m, n) &= \langle\langle \cos[m(\phi_1 - \phi_3) + n(\phi_2-\phi_4)]\rangle\rangle \nonumber \\
  \nonumber &- \langle\langle\cos[m(\phi_1-\phi_2)]\rangle\rangle\langle\langle\cos[n(\phi_1-\phi_2)]\rangle\rangle \label{eq:scmn}\\
  &\approx \langle v_m^2 v_n^2 \rangle - \langle v_m^2\rangle\langle v_n^2\rangle.
\end{align}
We show these model predictions in Fig.~\ref{fig:flow_corr} along with the \emph{normalized} symmetric cumulants
\begin{equation}
  \NSC(m,n) = \SC(m,n)/\langle v_m^2\rangle\langle v_n^2\rangle,
\end{equation}
which are expected to be less sensitive to the medium response and more sensitive to the properties of the initial state.
The solid lines are the MAP estimate of the present study, and the dashed lines are the MAP estimate of Ref.~\cite{Bernhard:2018hnz} which did not include nucleon substructure and was calibrated on Pb-Pb collisions at $\sqrts=2.76$ and 5.02~TeV.
We observe that the gap between $\SC(3,2)$ and $\SC(4,2)$ is generally wider in the present analysis than in Ref.~\cite{Bernhard:2018hnz}, as is the gap between the normalized symmetric cumulants $\NSC(3,2)$ and $\NSC(4,2)$.

@@
